---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# controlledburn 

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/controlledburn/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/controlledburn/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of controlledburn is to rasterize without materializing any pixel values.


This is an expression of the "cell abstraction" wish item [fasterize/issues/11](https://github.com/ecohealthalliance/fasterize/issues/11). 

- [x] name the package
- [ ] move to cpp11
- [ ] rasterize lines and points [fasterize/issues/30](https://github.com/ecohealthalliance/fasterize/issues/30)
- [ ] formats for import (wk, geos, grd, rct, triangles etc.)
- [ ] streaming with wkb/xy unpack with wk
- [ ] provide output options (see next)
- [x] copy logic from fasterize, and remove Armadillo array handling
- [x] remove use of raster objects, in favour of input extent and dimension
- [x] remove all trace of the raster package
- [x] implement return of the 'yline, xpix' and polygon ID info to user (see below)
- [x] make return of ylin,xpix structure efficient (CollectorList.h ftw)

## Outputs

Currently we get a list of triplets, so examples are unlist this to a 3-column matrix (and add 1). 

* two options, record presence of polygon OR ID of polygon
* a row-indexed (*yline*) set of edge instances (start, end *xpix*) along scanlines with the two options
* tools to format this meaningfully, and plot lazily (see example for quick plot)
* tools to materialize as actual raster data

I wanted this facility a long time, and tried to get discussion on it and tried
to implement it.  I also found this real world example, discussed in PROJ for
very fast lookup for large non-materialized (highly compressed) grids by Thomas
Knudsen:

https://github.com/OSGeo/PROJ/issues/1461#issuecomment-491501992


## Installation

You can install the development version of controlledburn like so:

``` r
remotes::install_github("hypertidy/controlledburn")
```

## Example

This is a basic example, this is fast, and shows that it works. See the todo list above. 

```{r example}
pols <- silicate::inlandwaters
library(vaster)
## define a raster (xmin, xmax, ymin, ymax), (ncol, nrow)
ext <- unlist(lapply(silicate::sc_vertex(pols), range))
dm <- c(500, 400)/10
r <- controlledburn:::laserize(pols, extent = ext,
                               dimension = dm, lines = T)

## our index is triplets of start,end,line where the polygon edge was detected - 
## this essentially an rle by scanline of start,end polygon coverage
index <- matrix(unlist(r, use.names = F), ncol = 3L, byrow = TRUE) + 1 ## plus one because 0-index internally
cr <- do.call(rbind, apply(index, 1, \(.x) cbind(seq(.x[1], .x[2]), .x[3])))
lcells_all <- sort(vaster::cell_from_row_col(dm, cr[,2], cr[,1]))

plot(silicate::PATH0(pols))
points(xy_from_cell(dm, ext, lcells_all), pch = ".")
str(index)

## convert to cells
cells <- vaster::cell_from_row_col(dm,  index[,c(3, 3)], index[,1:2])
str(cells)
```

It scales to very large tasks, less than a second for `250000x200000` dimension above, with 35Mb in memory output. 

The following is inefficient, but shows that we get the right result. 

```{r bad}
plot(vaster::xy_from_cell(dm, ext, cells))
## now go inefficient, this is every column,row index, then converted to cell, converted to xy
cr <- do.call(rbind, apply(index, 1, \(.x) cbind(seq(.x[1], .x[2]), .x[3])))
points(xy <- vaster::xy_from_cell(dm, ext, vaster::cell_from_row_col(dm, cr[,2], cr[,1])), pch = ".", col = "darkgrey")

r <- terra::rast(cbind(xy, 0), type = "xyz")

r[terra::cellFromXY(r, xy)] <- 1
terra::plot(r ,col = "firebrick")
plot(silicate::SC0(pols), add = TRUE)
```




## Code of Conduct
  
Please note that the controlledburn project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
