---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# controlledburn

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/controlledburn/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/controlledburn/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Rasterize geometry without materializing any pixel values. controlledburn
produces sparse tables for polygon, line, and point input — one type-pure
table per geometry kind.

For polygons: run-length-encoded interior cells and boundary cells with
exact partial coverage (`fraction` in [0, 1]). For lines: per-cell
absolute length in CRS units. For points: per-cell records with no
measure column.

The scanline algorithm is O(perimeter) in time and memory; no dense matrix
is allocated.

## Installation

```r
remotes::install_github("hypertidy/controlledburn")
```

## Usage

```{r usage}
library(controlledburn)
library(geos)

poly <- as_geos_geometry("POLYGON ((1 1, 9 1, 9 9, 1 9, 1 1))")

# Sparse output — no dense matrix allocated
r <- burn_scanline(poly, extent = c(0, 10, 0, 10), dimension = c(20L, 20L))
r

# Materialise only when you need it
mat <- materialise_chunk(r, c(0, 10, 0, 10))
```

### Default grid parameters

With no extent or dimension, controlledburn derives both from the geometry:

```{r helpers}
r <- burn_scanline(poly)
# extent from wk::wk_bbox(), 256 cells on the long axis
```

Or specify resolution:

```{r resolution}
r <- burn_scanline(poly, resolution = 0.5)
```

### Lines and points

Lines produce a `$lines` table with absolute length within each cell
(in CRS units, not a fraction):

```{r line-usage}
line <- as_geos_geometry("LINESTRING (0 5, 10 5)")
r <- burn_scanline(line, extent = c(0, 10, 0, 10), dimension = c(20L, 20L))
r$lines
```

Points produce a `$points` table with one record per cell hit (no
measure column — a point is either in a cell or it isn't):

```{r point-usage}
pts <- as_geos_geometry(c("POINT (2 3)", "POINT (7 8)"))
r <- burn_scanline(pts, extent = c(0, 10, 0, 10), dimension = c(20L, 20L))
r$points
```

### Geometry input

`burn_scanline()` accepts `geos_geometry`, `sfc` (sf), `wk::wkb()`,
`blob`, or a list of raw WKB vectors. The raw-WKB path is compatible
with `vapour::vapour_read_geometry()` and `gdalraster::GDALVector`
output. For `terra::vect()` input, round-trip via
`geos::as_geos_geometry()`.

### Shared boundary complementarity

Adjacent polygons with shared edges produce complementary coverage fractions
that sum to exactly 1.0 in every boundary cell — no gaps, no overlaps:

```{r shared}
left  <- as_geos_geometry("POLYGON ((0 0, 5 0, 5 10, 0 10, 0 0))")
right <- as_geos_geometry("POLYGON ((5 0, 10 0, 10 10, 5 10, 5 0))")

r <- burn_scanline(c(left, right), extent = c(0,10,0,10), dimension = c(20L,20L))

# Coverage sums to 1.0 in every touched cell
mat1 <- materialise_chunk(r, id = 1)
mat2 <- materialise_chunk(r, id = 2)
max(mat1 + mat2)
#> [1] 1
```

## Output format

`burn_scanline()` returns a list with class `"controlledburn"`:

- **`runs`**: `data.frame(row, col_start, col_end, id)` — polygon interior
  cells (full coverage), run-length encoded by row.
- **`edges`**: `data.frame(row, col, fraction, id)` — polygon boundary
  cells with partial coverage; `fraction` is in (0, 1).
- **`lines`**: `data.frame(row, col, length, id)` — line cells; `length`
  is the absolute length of the line within the cell, in CRS units.
- **`points`**: `data.frame(row, col, id)` — point cells; no measure
  column (a point is either in a cell or it isn't).
- **`extent`**: `c(xmin, xmax, ymin, ymax)`
- **`dimension`**: `c(ncol, nrow)`

Tables are populated for whichever geometry kinds are in the input. Each
table's measure column means exactly one thing — `$edges$fraction` is
dimensionless, `$lines$length` is in CRS units, points have no measure.
This separation is deliberate: the three measures are different
mathematical objects and combining them in one column would silently
mix units.

`burn_sparse()` is polygon-only and returns just `$runs` and `$edges`.
Line and point input there is rejected with an error pointing at
`burn_scanline()`.

This is the natural output of scanline rasterization — no dense matrix is
allocated until `materialise_chunk()` is called.

## Performance

Scanline algorithm scales with perimeter, not area. The comparison
table is polygon-only — `burn_sparse()` is the older bbox-bounded
exactextract path and does not accept line or point input:

| Shape | Resolution | Scanline | Dense (burn_sparse) | Speedup |
|-------|-----------|----------|-------------------|---------|
| Star | 3200×3200 | 13 ms | 225 ms | 17× |
| Jagged | 3200×3200 | 15 ms | 136 ms | 9× |
| NC counties | 2000×800 | 29 ms | 61 ms | 2× |

Memory for real-world grids (CGAZ at 32K×16K, ~500M cells): ~50 MB sparse
vs ~2 GB dense.

## Real-world example: global administrative boundaries

The CGAZ (Geo Boundaries) ADM0 dataset is roughly <!-- TODO: a few hundred MB
of WKB across ~250 polygons of varying complexity --> a useful stress test
for both shape complexity and scale.

```{r real-world-fetch}
v <- new(gdalraster::GDALVector, "/vsicurl/https://github.com/mdsumner/geoboundaries/releases/download/latest/geoBoundariesCGAZ_ADM0.parquet")
v$returnGeomAs ## WKB
gcol <- v$getGeometryColumn()
v$setIgnoredFields( setdiff(v$getFieldNames(), gcol))
wkbgeom <- wk::wkb(v$fetch(-1)[[gcol]])
v$close()

system.time(burn_scanline(wkbgeom))

system.time(r1 <- burn_scanline(wkbgeom, dimension = c(8192, 4096)))
str(r1)
```

<!-- TODO: short observation about what the timing demonstrates —
e.g. "X seconds for the full world at 8K×4K, scaling linearly with
perimeter as the resolution increases." -->

```{r real-world-scaled}
system.time(r1 <- burn_scanline(wkbgeom, dimension = c(8192, 4096) * 20))
pryr::object_size(r1)
tibble::as_tibble(r1$runs)
tibble::as_tibble(r1$edges)
r1[c("extent", "dimension")]
```


## History

controlledburn was derived from
[fasterize](https://github.com/ecohealthalliance/fasterize) by Noam
Ross (EcoHealth Alliance), removing Armadillo and raster package
dependencies. See [NEWS](NEWS.md) for the version history; the design
record lives in `inst/docs-design/`.

## See also

- [vaster](https://github.com/hypertidy/vaster) — primitive grid
  cell ↔ xy operations; consumes the `(row, col, …)` schema this
  package emits.
- [silicate](https://github.com/hypertidy/silicate) — the
  primitives-first geometry stance this package follows
  (segments and vertices as first-class objects).
- [polymer2](https://github.com/hypertidy/polymer2) — sparse
  geometry overlay; consumes the `(row, col, fraction, id)` schema
  this package emits for polygons.
- [exactextractr](https://github.com/isciences/exactextractr) —
  raster extraction with polygon weights; the source of the
  exactextract C++ algorithm vendored here.
- <!-- TODO: any other neighbours worth flagging for line/point
  consumers — `lazysf`, `wk`, `geos`, `gdalraster`? -->

## Code of Conduct

Please note that the controlledburn project is released with a
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
