---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# controlledburn

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/controlledburn/actions/workflows/R-CMD-check-check.yaml/badge.svg)](https://github.com/hypertidy/controlledburn/actions/workflows/R-CMD-check-check.yaml)
<!-- badges: end -->

Rasterize geometry without materializing any pixel values. controlledburn
produces sparse tables for polygon, line, and point input — one type-pure
table per geometry kind.

Two modes: **coverage** (exact analytical coverage fractions via vendored
exactextract math) and **approx** (cell-centre rule, fasterize
semantics, runs only). Both are O(perimeter) in time and memory; no
dense matrix is allocated.

Dependencies: `cpp11`, `wk`. No GEOS, no sf, no Armadillo.

## Installation

```r
remotes::install_github("hypertidy/controlledburn")
```

## Usage

```{r usage}
library(controlledburn)
library(geos)

poly <- as_geos_geometry("POLYGON ((1 1, 8.9 1, 8.9 9.12, 1 9.12, 1 1))")

# Coverage mode (default): exact boundary fractions
r <- burn(poly, extent = c(0, 10, 0, 10), dimension = c(20L, 20L))
r

# Approx mode: cell-centre rule, runs only, no edges
r_approx <- burn(poly, extent = c(0, 10, 0, 10), dimension = c(20L, 20L),
                 mode = "approx")
r_approx

# Materialize only when you need it
mat <- materialize_chunk(r)
ximage::ximage(mat, extent = c(0, 10, 0, 10), asp = 1)
plot(poly, add = TRUE)
unique(c(mat))
## no variation in the approx mode
unique(c(materialise_chunk(r_approx)))
```

### Default grid parameters

With no extent or dimension, controlledburn derives both from the geometry:

```{r helpers}
r <- burn(poly)
# extent from wk::wk_bbox(), 256 cells on the long axis
```

Or specify resolution:

```{r resolution}
r <- burn(poly, resolution = 0.5)
```

### Lines and points

Lines produce a `$lines` table with absolute length within each cell
(in CRS units, not a fraction):

```{r line-usage}
line <- as_geos_geometry("LINESTRING (0 5, 10 5)")
r <- burn(line, extent = c(0, 10, 0, 10), dimension = c(20L, 20L))
head(r$lines)
```

Points produce a `$points` table with one record per cell hit (no
measure column — a point is either in a cell or it isn't):

```{r point-usage}
pts <- as_geos_geometry(c("POINT (2 3)", "POINT (7 8)"))
r <- burn(pts, extent = c(0, 10, 0, 10), dimension = c(20L, 20L))
r$points
```

### Geometry input

`burn()` accepts `geos_geometry`, `sfc` (sf), `wk::wkb()`,
`blob`, or a list of raw WKB vectors. The raw-WKB path is compatible
with `vapour::vapour_read_geometry()` and `gdalraster::GDALVector`
output.

controlledburn rasterizes whatever geometry it's given. Self-intersecting
rings, unclosed polygons, repeated vertices, near-degenerate inputs — all
go through the same fast path. If your input has topological issues that
matter for your science, you'll see it in the output and can decide what
to do. If they don't matter, you've saved the cost of validating them.
Either way the package trusts your judgement on what valid means in
context. For when you do want to check or repair,
`geos::geos_is_valid()` and `geos::geos_make_valid()` are perfectly
suitable.

### Shared boundary complementarity

Adjacent polygons with shared edges produce complementary coverage fractions
that sum to exactly 1.0 in every boundary cell — no gaps, no overlaps:

```{r shared}
left  <- as_geos_geometry("POLYGON ((0 0, 5 0, 5 10, 0 10, 0 0))")
right <- as_geos_geometry("POLYGON ((5 0, 10 0, 10 10, 5 10, 5 0))")

r <- burn(c(left, right), extent = c(0, 10, 0, 10), dimension = c(20L, 20L))

# Coverage sums to 1.0 in every touched cell
mat <- materialize_chunk(r)
max(mat)
```

## Output format

`burn()` returns a list with class `"controlledburn"`:

- **`runs`**: `data.frame(row, col_start, col_end, id)` — polygon interior
  cells (full coverage), run-length encoded by row. In approx mode,
  boundary cells classified as "inside" also appear here.
- **`edges`**: `data.frame(row, col, fraction, id)` — polygon boundary
  cells with partial coverage; `fraction` is in (0, 1). Empty in approx
  mode.
- **`lines`**: `data.frame(row, col, length, id)` — line cells; `length`
  is the absolute length of the line within the cell, in CRS units.
- **`points`**: `data.frame(row, col, id)` — point cells; no measure
  column (a point is either in a cell or it isn't).
- **`extent`**: `c(xmin, xmax, ymin, ymax)`
- **`dimension`**: `c(ncol, nrow)`

Tables are populated for whichever geometry kinds are in the input. Each
table's measure column means exactly one thing — `$edges$fraction` is
dimensionless, `$lines$length` is in CRS units, points have no measure.
This separation is deliberate: the three measures are different
mathematical objects and combining them in one column would silently
mix units.

This is the natural output of scanline rasterization — no dense matrix is
allocated until `materialize_chunk()` is called.

## Performance

### controlledburn vs fasterize

Benchmarked on CGAZ (218 country polygons, 10.1M vertices). Approx mode
uses a lightweight edge-row intersection sweep that bypasses the
exactextract walker entirely.

| Grid | Cells | cb approx | fasterize | Winner |
|------|-------|-----------|-----------|--------|
| 256 × 128 | 33K | 1.0s | 0.5s | fasterize |
| 4096 × 2048 | 8.4M | 1.1s | 0.4s | fasterize |
| **16384 × 8192** | **134M** | **1.2s** | **2.8s** | **cb** |
| 32768 × 16384 | 537M | 1.4s | 9.7s | cb |
| 65536 × 32768 | 2.1B | 1.7s | OOM | cb only |
| 131072 × 65536 | 8.6B | 2.4s | OOM | cb only |

Crossover at ~134 million cells. Above that, fasterize's dense raster
allocation dominates. At 8.6 billion cells, controlledburn completes
in 2.4 seconds where fasterize cannot allocate.

On NC counties at 2000×800, approx mode produces cell-for-cell identical
output to fasterize (zero discrepant cells).

### Extreme scale

Antarctic rock outcrop polygons (25,954 geometries) against a REMA 2m DEM
grid (2.7 million × 2.9 million pixels, ~8 trillion cells): 8 seconds,
438 MB of sparse output, 99.8% sparsity. A dense raster at this
resolution would require ~30 TB.

## CGAZ example

```{r cgaz-example}
g <- geos::as_geos_geometry(wk::wkb(vapour::vapour_read_geometry(sds::CGAZ())))

system.time(r <- burn(g, dimension = c(2560L, 1280L), mode = "approx"))
r
```

## History

controlledburn was derived from
[fasterize](https://github.com/ecohealthalliance/fasterize) by Noam
Ross (EcoHealth Alliance). The exact coverage fraction algorithm is from
Daniel Baston's [exactextract](https://github.com/isciences/exactextract)
C++ library, vendored as 9 GEOS-free analytical geometry files.

The development history: fasterize's scanline algorithm → sparse
run-length output → exactextract integration for exact coverage
fractions → native WKB parser and ring walker replacing GEOS → unified
polygon/line/point rasterization → dual-mode engine (coverage + approx)
→ lightweight approx sweep → GEOS dependency removed entirely.

See `vignette("architecture")` for the full story, [NEWS](NEWS.md) for
the version history, and `inst/docs-design/` for design records.

## See also

- [vaster](https://github.com/hypertidy/vaster) — primitive grid
  cell ↔ xy operations; consumes the `(row, col, …)` schema this
  package emits.
- [silicate](https://github.com/hypertidy/silicate) — the
  primitives-first geometry stance this package follows
  (segments and vertices as first-class objects).
- [polymer2](https://github.com/hypertidy/polymer2) — sparse
  geometry overlay; consumes the `(row, col, fraction, id)` schema
  this package emits for polygons.
- [exactextractr](https://github.com/isciences/exactextractr) —
  raster extraction with polygon weights; the source of the
  exactextract C++ algorithm vendored here.

## Code of Conduct

Please note that the controlledburn project is released with a
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
