---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# laserize 

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/laserize/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/laserize/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of laserize is to rasterize without materializing any pixel values.

WARNING: does not do anything yet, it's just me learning C++. Do not use, VERY VERY WIP


This is an expression of my WISHUE "cell abstraction" [fasterize/issues/11](https://github.com/ecohealthalliance/fasterize/issues/11). 


- [x] copy logic from fasterize, and remove Armadillo array handling
- [x] remove use of raster objects, in favour of input extent and dimension
- [x] remove all trace of the raster package
- [x] implement return of the 'yline, xpix' and polygon ID info to user (see below)
- [ ] make return of ylin,xpix structure efficient (currently growing the vector)
- [ ] move to cpp11
- [ ] rasterize lines and points [fasterize/issues/30](https://github.com/ecohealthalliance/fasterize/issues/30)
- [ ] consider formats other than sf (wk, geos, grd, rct, triangles etc.)

## Outputs

* two options, record presence of polygon OR ID of polygon
* a row-indexed (*yline*) set of edge instances (start, end *xpix*) along scanlines with the two options
* tools to format this meaningfully, and plot lazily
* tools to materialize as actual raster data

For the record, I wanted this facility before I read this issue - but here's a real world example, discussed in PROJ for
very fast lookup for large non-materialized (highly compressed) grids by Thomas Knudsen: 

https://github.com/OSGeo/PROJ/issues/1461#issuecomment-491501992


## Installation

You can install the development version of laserize like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

## Example

This is a basic example, currently the CPP code is slow because it grows the output vectors, and the matrix orientation
used below is not the intended end point. But, just shows this works. See the todo list above. 

```{r example}
pdata <-
  structure(list(sfg_id = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3),
                 polygon_id = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3),
                 linestring_id = c(1, 1, 1, 1, 1, 2, 2, 2, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1),
                 x = c(-180, -140, 10, -140, -180, -150, -100, -110, -150, -10, 140, 160, 140, -10, -125, 0, 40, 15, -125),
                 y = c(-20, 55, 0, -60, -20, -20, -10, 20, -20, 0, 60, 0, -55, 0, 0, 60, 5, -45, 0)),
            class = "data.frame", row.names = c(NA, 19L))

pols <- sfheaders::sf_polygon(pdata, x= "x", y = "y", polygon_id = "polygon_id", linestring_id = "linestring_id")
## define a raster (xmin, xmax, ymin, ymax), (ncol, nrow)
ext <- c(range(pdata$x), range(pdata$y))
dm <- c(50, 40)
r <- laserize:::laserize(pols, extent = ext,
                               dimension = dm)

## our index is pairs of y, x where the polygon edge was detected - 
## this essentially an rle by scanline of start,end polygon coverage
index <- matrix(r, ncol = 3L, byrow = TRUE)

str(index)
## see tests for current code, there's still investigation of an out by one thing 
```




## Code of Conduct
  
Please note that the laserize project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
