---
title: "Algorithms: scanline fill, exact coverage, and the sparse intersection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Algorithms: scanline fill, exact coverage, and the sparse intersection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 5,
  out.width = "80%",
  dpi = 120
)
library(ggplot2)
library(controlledburn)

# Shared polygon for all illustrations
poly_x <- c(1.3, 6.7, 5.2, 2.8, 1.3)
poly_y <- c(1.5, 2.3, 5.8, 4.2, 1.5)
nx <- 8; ny <- 7

# Point-in-polygon via winding number
pip <- function(px, py, poly_x, poly_y) {
  n <- length(poly_x) - 1
  winding <- 0
  for (i in seq_len(n)) {
    y0 <- poly_y[i]; y1 <- poly_y[i + 1]
    x0 <- poly_x[i]; x1 <- poly_x[i + 1]
    if (y0 <= py) {
      if (y1 > py) {
        cross_x <- x0 + (py - y0) / (y1 - y0) * (x1 - x0)
        if (px < cross_x) winding <- winding + 1
      }
    } else {
      if (y1 <= py) {
        cross_x <- x0 + (py - y0) / (y1 - y0) * (x1 - x0)
        if (px < cross_x) winding <- winding - 1
      }
    }
  }
  winding != 0
}

# Edge-scanline intersections
edge_crossings <- function(poly_x, poly_y, y_mid) {
  n <- length(poly_x) - 1
  crossings <- numeric(0)
  for (i in seq_len(n)) {
    y0 <- poly_y[i]; y1 <- poly_y[i + 1]
    x0 <- poly_x[i]; x1 <- poly_x[i + 1]
    if ((y0 < y_mid & y1 >= y_mid) | (y1 < y_mid & y0 >= y_mid)) {
      t <- (y_mid - y0) / (y1 - y0)
      crossings <- c(crossings, x0 + t * (x1 - x0))
    }
  }
  sort(crossings)
}

# Base grid theme
theme_grid <- function() {
  theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank())
}
```

## The shared question

Three related tools — fasterize, exactextract, and controlledburn — all
answer the same underlying question:

> **How does this geometry relate to this grid?**

They differ in *what they do with the answer*. Understanding the
algorithms means understanding both the geometric computation and the
motivation that shaped each tool's design.

## The scanline algorithm

At the heart of all polygon rasterization is the **scanline fill**: walk
a horizontal line across each row of the grid, find where polygon edges
cross it, and fill between the crossings.

```{r scanline-fill, fig.cap = "Scanline fill: for each row, polygon edges are intersected with the row's horizontal scanline (dashed blue). Red dots mark the intersections. Between each pair of crossings, all cells along the blue segment are 'inside' the polygon."}
scanline_segs <- do.call(rbind, lapply(seq(0.5, ny - 0.5), function(y_mid) {
  xints <- edge_crossings(poly_x, poly_y, y_mid)
  if (length(xints) >= 2) {
    pairs <- matrix(xints, ncol = 2, byrow = TRUE)
    data.frame(y = y_mid, xstart = pairs[, 1], xend = pairs[, 2])
  }
}))

scanline_pts <- do.call(rbind, lapply(seq(0.5, ny - 0.5), function(y_mid) {
  xints <- edge_crossings(poly_x, poly_y, y_mid)
  if (length(xints) >= 2) data.frame(y = y_mid, x = xints)
}))

ggplot() +
  geom_vline(xintercept = 0:nx, colour = "grey85", linewidth = 0.3) +
  geom_hline(yintercept = 0:ny, colour = "grey85", linewidth = 0.3) +
  geom_hline(yintercept = seq(0.5, ny - 0.5), colour = "steelblue",
             linewidth = 0.2, linetype = "dashed") +
  geom_polygon(aes(x = poly_x, y = poly_y),
               fill = NA, colour = "black", linewidth = 0.8) +
  geom_segment(data = scanline_segs,
               aes(x = xstart, xend = xend, y = y, yend = y),
               colour = "steelblue", linewidth = 1.5, alpha = 0.5) +
  geom_point(data = scanline_pts, aes(x, y),
             colour = "firebrick", size = 2.5) +
  coord_equal(xlim = c(0, nx), ylim = c(0, ny)) +
  theme_grid()
```

This is efficient: the cost is proportional to the number of edges times
the number of rows they span — **O(perimeter)**, not O(area). A small
polygon on a billion-cell grid produces a handful of crossings per row.

The question is: what do you *do* with those crossings?

## Two answers to the boundary question

The crossings tell you which cells are interior (fully between edge
pairs). But what about the cells where the edge actually crosses? A
polygon edge slices through a cell, leaving part inside and part outside.
The two fundamental approaches:

### Cell-centre rule (fasterize / approx mode)

The simplest answer: a cell is "in" if its **centre point** falls inside
the polygon. Binary, no fractions. Fast to compute — you already know
the crossing x-coordinates, so you just check whether the cell centre
is between a pair.

```{r center-rule, fig.cap = "Cell-centre rule: each cell is classified by whether its centre point (dot) falls inside the polygon. Red = inside, grey = outside. No partial coverage — every cell is either fully in or fully out."}
centers <- expand.grid(
  cx = seq(0.5, nx - 0.5, by = 1),
  cy = seq(0.5, ny - 0.5, by = 1)
)
centers$inside <- mapply(pip, centers$cx, centers$cy,
                         MoreArgs = list(poly_x = poly_x, poly_y = poly_y))

# Highlight "in" cells
in_rects <- centers[centers$inside, ]
in_rects <- data.frame(
  xmin = in_rects$cx - 0.5, xmax = in_rects$cx + 0.5,
  ymin = in_rects$cy - 0.5, ymax = in_rects$cy + 0.5
)

ggplot() +
  geom_rect(data = in_rects,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "steelblue", alpha = 0.4) +
  geom_vline(xintercept = 0:nx, colour = "grey80", linewidth = 0.3) +
  geom_hline(yintercept = 0:ny, colour = "grey80", linewidth = 0.3) +
  geom_polygon(aes(x = poly_x, y = poly_y),
               fill = NA, colour = "black", linewidth = 0.8) +
  geom_point(data = centers, aes(cx, cy, colour = inside), size = 1.8) +
  scale_colour_manual(values = c("TRUE" = "firebrick", "FALSE" = "grey60"),
                      guide = "none") +
  coord_equal(xlim = c(0, nx), ylim = c(0, ny)) +
  theme_grid()
```

This is what **fasterize** does, and what controlledburn's **approx
mode** reproduces. It's fast, simple, and correct for many use cases —
but it doesn't conserve area. The total burned area (count of cells ×
cell area) won't match the polygon's true area.

### Exact coverage fractions (exactextract / coverage mode)

The precise answer: for each boundary cell, compute the **exact fraction**
of the cell that falls inside the polygon. Interior cells get fraction
1.0, exterior get 0.0, and boundary cells get an analytically computed
value between 0 and 1.

```{r exact-coverage, fig.cap = "Exact coverage: interior cells (blue, fraction = 1.0) are fully inside. Boundary cells (coral) have an exact coverage fraction — the proportion of the cell area that falls inside the polygon edge. Total coverage equals the polygon's true area."}
# Classify cells
cell_class <- function(cx, cy) {
  xmin <- cx - 0.5; xmax <- cx + 0.5
  ymin <- cy - 0.5; ymax <- cy + 0.5
  n <- length(poly_x) - 1
  edge_crosses <- FALSE
  for (i in seq_len(n)) {
    x0 <- poly_x[i]; y0 <- poly_y[i]
    x1 <- poly_x[i+1]; y1 <- poly_y[i+1]
    if (max(x0, x1) >= xmin & min(x0, x1) <= xmax &
        max(y0, y1) >= ymin & min(y0, y1) <= ymax) {
      edge_crosses <- TRUE; break
    }
  }
  if (!edge_crosses) {
    return(if (pip(cx, cy, poly_x, poly_y)) "interior" else "exterior")
  }
  "boundary"
}

centers$class <- mapply(cell_class, centers$cx, centers$cy)

cell_rects <- data.frame(
  xmin = centers$cx - 0.5, xmax = centers$cx + 0.5,
  ymin = centers$cy - 0.5, ymax = centers$cy + 0.5,
  class = centers$class
)

ggplot() +
  geom_rect(data = cell_rects[cell_rects$class != "exterior", ],
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
                fill = class),
            alpha = 0.5, colour = "grey70", linewidth = 0.2) +
  geom_vline(xintercept = 0:nx, colour = "grey80", linewidth = 0.3) +
  geom_hline(yintercept = 0:ny, colour = "grey80", linewidth = 0.3) +
  geom_polygon(aes(x = poly_x, y = poly_y),
               fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_manual(values = c("interior" = "steelblue",
                                "boundary" = "coral"),
                    labels = c("interior" = "interior (fraction = 1.0)",
                               "boundary" = "boundary (0 < fraction < 1)")) +
  coord_equal(xlim = c(0, nx), ylim = c(0, ny)) +
  theme_grid()
```

This is what Daniel Baston's **exactextract** computes, and what
controlledburn's **coverage mode** uses. The analytical traversal walks
each polygon edge through the grid, computing the exact sub-pixel area
on each side of the edge in every cell it crosses. The total coverage
(sum of all fractions × cell area) equals the polygon's true geometric
area.

## The same polygon, two classifications

The difference is only at the boundary. Both approaches agree on
interior cells (fully inside) and exterior cells (fully outside). They
disagree on the ~O(perimeter) boundary cells:

```{r side-by-side, fig.width = 9, fig.height = 4.5, fig.cap = "The same polygon, two classification schemes. Left: cell-centre rule — every cell is in or out. Right: exact coverage — boundary cells (coral) carry a fractional weight. Interior cells (blue) are the same in both."}
# Side-by-side
centers$approx <- ifelse(centers$inside, "in", "out")
centers$coverage <- centers$class

d1 <- data.frame(
  xmin = centers$cx - 0.5, xmax = centers$cx + 0.5,
  ymin = centers$cy - 0.5, ymax = centers$cy + 0.5,
  fill = ifelse(centers$approx == "in", "approx: in", "none"),
  panel = "Cell-centre rule (approx)"
)
d2 <- data.frame(
  xmin = centers$cx - 0.5, xmax = centers$cx + 0.5,
  ymin = centers$cy - 0.5, ymax = centers$cy + 0.5,
  fill = ifelse(centers$coverage == "interior", "interior",
                ifelse(centers$coverage == "boundary", "boundary", "none")),
  panel = "Exact coverage (coverage)"
)

dd <- rbind(d1, d2)
dd <- dd[dd$fill != "none", ]

poly_df <- data.frame(x = poly_x, y = poly_y)
grid_v <- data.frame(x = 0:nx)
grid_h <- data.frame(y = 0:ny)

ggplot(dd) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
                fill = fill),
            alpha = 0.5, colour = "grey70", linewidth = 0.2) +
  geom_vline(data = grid_v, aes(xintercept = x),
             colour = "grey80", linewidth = 0.3) +
  geom_hline(data = grid_h, aes(yintercept = y),
             colour = "grey80", linewidth = 0.3) +
  geom_polygon(data = poly_df, aes(x, y),
               fill = NA, colour = "black", linewidth = 0.8) +
  scale_fill_manual(
    values = c("approx: in" = "steelblue",
               "interior" = "steelblue",
               "boundary" = "coral"),
    labels = c("approx: in" = "cell centre inside",
               "interior" = "interior (1.0)",
               "boundary" = "boundary (fractional)")
  ) +
  facet_wrap(~panel) +
  coord_equal(xlim = c(0, nx), ylim = c(0, ny)) +
  theme_grid() +
  theme(strip.text = element_text(face = "bold", size = 11))
```

## Three motivational frames

The algorithm determines what you can compute. But the *motivation*
determines which algorithm you reach for.

### fasterize: "which value goes in this pixel?"

fasterize was built for raster map production. The typical workflow:
assign a value from each polygon's attribute table to the cells it
covers. Three knobs control the output:

- **mask**: burn 1 for every polygon. Binary presence/absence.
- **field**: copy a column value (e.g. land-use class, country ID) into
  each cell.
- **fun**: when polygons overlap, resolve the collision — `"last"` (most
  recent polygon wins), `"sum"` (accumulate), `"min"`, `"max"`, etc.

The cell-centre rule is perfect here: you need a single winner per
cell, and fractional coverage doesn't help choose which polygon "owns"
a pixel.

### exactextract: "what's the weighted aggregate under this polygon?"

exactextract was built for zonal statistics. The typical workflow: given
a raster of (say) temperature values and a set of polygons, compute the
area-weighted mean temperature in each polygon. The coverage fraction is
the weight:

$$\bar{T}_{\text{polygon}} = \frac{\sum_i f_i \cdot T_i}{\sum_i f_i}$$

where $f_i$ is the coverage fraction and $T_i$ is the raster value in
cell $i$. Without exact fractions, boundary cells bias the result —
especially for small polygons where boundary cells dominate.

### controlledburn: "give me the geometry–grid relationship"

controlledburn separates the geometric computation from the
application. It produces the **sparse intersection table** — runs,
edges, fractions — and lets the caller decide what to do with it:

- Use the fractions as weights for zonal stats (like exactextract)
- Use the runs as a mask (like fasterize with `fun = "last"`)
- Use the `id` column to track which polygon covers which cell
- Use the runs at extreme resolution where no dense raster could be
  allocated
- Use the line lengths or point hits for non-polygon operations

The sparse representation makes this possible at scales where
materializing a dense raster is infeasible. The 8-trillion-cell
Antarctic example in the README produces 438 MB of sparse runs from
a 2.7-million × 2.9-million pixel grid — a dense raster at that
resolution would require ~30 TB.

## controlledburn's two code paths

### Approx mode: the lightweight sweep

Approx mode reimplements the scanline fill as a direct **edge–row
intersection sweep**. For each polygon edge, it computes the
x-coordinate where the edge crosses each row's scanline, accumulates
winding numbers, and fills between crossings. ~120 lines of C++, no
traversal bookkeeping.

```{r approx-detail, fig.cap = "Approx mode in detail: for each row, polygon edges are intersected with the scanline at the cell-centre y-coordinate. The winding number accumulates left-to-right: it goes nonzero when entering the polygon (green arrow) and returns to zero when leaving (red arrow). Cells whose centre falls in a nonzero-winding span are classified as 'inside' and emitted as runs."}
# Illustration of winding accumulation for one row
y_show <- 3.5
xints <- edge_crossings(poly_x, poly_y, y_show)

# Build winding annotation
winding_df <- data.frame(
  x = c(0, xints),
  xend = c(xints, nx),
  y = y_show, 
  winding = cumsum(c(0, rep(c(1, -1), length.out = length(xints))))
)

ggplot() +
  geom_vline(xintercept = 0:nx, colour = "grey85", linewidth = 0.3) +
  geom_hline(yintercept = 0:ny, colour = "grey85", linewidth = 0.3) +
  # Highlight the row
  geom_rect(aes(xmin = 0, xmax = nx, ymin = y_show - 0.5, ymax = y_show + 0.5),
            fill = "lightyellow", alpha = 0.5) +
  # Polygon
  geom_polygon(aes(x = poly_x, y = poly_y),
               fill = NA, colour = "black", linewidth = 0.8) +
  # Winding spans
  geom_segment(data = winding_df[winding_df$winding != 0, ],
               aes(x = x, xend = xend, y = y, yend = y),
               colour = "steelblue", linewidth = 2, alpha = 0.4) +
  # Crossing points with arrows
  geom_point(data = data.frame(x = xints, y = y_show),
             aes(x, y), colour = "firebrick", size = 3) +
  # Winding labels
  geom_text(data = winding_df,
            aes(x = (x + xend) / 2, y = y + 0.35,
                label = paste0("w=", winding)),
            size = 3.5, colour = "grey30") +
  # Scanline
  geom_hline(yintercept = y_show, colour = "steelblue",
             linewidth = 0.3, linetype = "dashed") +
  coord_equal(xlim = c(0, nx), ylim = c(0, ny)) +
  theme_grid() +
  labs(title = sprintf("Winding number sweep at y = %.1f", y_show))
```

The boundary convention is **left-inclusive**: when an edge crossing
falls exactly on a cell centre, the cell is classified as "inside".
This matches fasterize's behaviour — on constructed geometries (aligned
rectangles, offset rectangles, triangles, holes), approx mode produces
cell-for-cell identical output to fasterize.

### Coverage mode: the exactextract walker

Coverage mode uses the full traversal engine from exactextract. For each
polygon ring, it walks the ring through the grid cell-by-cell,
recording the exact coordinates where the ring enters and exits each
cell. From these traversal records, it computes the analytically exact
fraction of each boundary cell that falls inside the polygon.

```{r coverage-detail, fig.cap = "Coverage mode boundary cell detail. The polygon edge enters a cell on one side and exits on another. The traversal coordinates define a sub-polygon within the cell whose area is the coverage fraction. This is computed analytically — no sampling, no approximation."}
# Zoom into a single boundary cell to show the traversal
# Pick cell (3, 3) — center (2.5, 2.5)
cell_cx <- 2.5; cell_cy <- 2.5
cell_xmin <- 2; cell_xmax <- 3; cell_ymin <- 2; cell_ymax <- 3

# Find where polygon edges cross this cell
# Edge from (1.3, 1.5) to (6.7, 2.3) crosses this cell
# Entry: left side at x=2, y = 1.5 + (2-1.3)/(6.7-1.3) * (2.3-1.5) = 1.5 + 0.1296*0.8 = 1.604
# But that's below the cell... let me find the right edge
# Edge from (2.8, 4.2) to (1.3, 1.5): crosses this cell
# At y=3: x = 2.8 + (3-4.2)/(1.5-4.2) * (1.3-2.8) = 2.8 + 0.4444*(-1.5) = 2.133
# At y=2: x = 2.8 + (2-4.2)/(1.5-4.2) * (1.3-2.8) = 2.8 + 0.8148*(-1.5) = 1.578

# Let me use a cell that's easier to illustrate
# Cell (5, 3) — center (4.5, 2.5), edge from (1.3,1.5) to (6.7,2.3) crosses bottom
cell_cx <- 4.5; cell_cy <- 2.5
cell_xmin <- 4; cell_xmax <- 5; cell_ymin <- 2; cell_ymax <- 3

# Edge 1: (1.3, 1.5) -> (6.7, 2.3)
# At x=4: y = 1.5 + (4-1.3)/(6.7-1.3) * (2.3-1.5) = 1.5 + 0.5 * 0.8 = 1.9
# At x=5: y = 1.5 + (5-1.3)/(6.7-1.3) * (2.3-1.5) = 1.5 + 0.685 * 0.8 = 2.048
# So this edge enters at (4, 1.9) — below the cell. Enters cell bottom at y=2:
# x at y=2: x = 1.3 + (2-1.5)/(2.3-1.5) * (6.7-1.3) = 1.3 + 0.625 * 5.4 = 4.675
# Exits at x=5: y = 2.048
# So the traversal within this cell is (4.675, 2) -> (5, 2.048)

# The polygon coverage in this cell: everything above the line from 
# (4.675, 2) to (5, 2.048), clipped to the cell
# Let's build a simple illustration

entry <- c(4.675, 2.0)
exit  <- c(5.0, 2.048)

# The covered region (below the top of cell, above the edge, within cell walls)
# It's the cell minus the triangle below the edge
covered_poly <- data.frame(
  x = c(cell_xmin, entry[1], exit[1], cell_xmax, cell_xmax, cell_xmin),
  y = c(cell_ymin, entry[2], exit[2], exit[2], cell_ymax, cell_ymax)
)

# Actually let me simplify — show the edge cutting through the cell
# and shade the "inside" portion
# For this edge going up-right, "inside" (polygon interior) is above/left

ggplot() +
  # Cell boundary
  geom_rect(aes(xmin = cell_xmin, xmax = cell_xmax,
                ymin = cell_ymin, ymax = cell_ymax),
            fill = "grey95", colour = "black", linewidth = 0.8) +
  # Covered area
  geom_polygon(data = covered_poly, aes(x, y),
               fill = "steelblue", alpha = 0.3) +
  # Edge traversal through cell
  geom_segment(aes(x = entry[1] - 0.3, xend = exit[1] + 0.3,
                   y = entry[2] - 0.3 * (exit[2]-entry[2])/(exit[1]-entry[1]),
                   yend = exit[2] + 0.3 * (exit[2]-entry[2])/(exit[1]-entry[1])),
               colour = "black", linewidth = 0.8) +
  # Entry and exit points
  geom_point(aes(x = c(entry[1], exit[1]), y = c(entry[2], exit[2])),
             colour = "firebrick", size = 3) +
  # Labels
  annotate("text", x = 4.3, y = 2.7, label = "inside\n(covered)",
           colour = "steelblue", fontface = "bold", size = 4) +
  annotate("text", x = 4.85, y = 2.15, label = "outside",
           colour = "grey50", size = 3.5) +
  annotate("text", x = entry[1], y = entry[2] - 0.08, label = "entry",
           colour = "firebrick", size = 3, vjust = 1) +
  annotate("text", x = exit[1] + 0.02, y = exit[2] + 0.05, label = "exit",
           colour = "firebrick", size = 3, hjust = 0) +
  coord_equal(xlim = c(cell_xmin - 0.2, cell_xmax + 0.2),
              ylim = c(cell_ymin - 0.2, cell_ymax + 0.2)) +
  theme_grid() +
  labs(title = "Boundary cell: coverage fraction from edge traversal",
       subtitle = sprintf("fraction = %.3f (shaded area / cell area)",
                          1 - (0.5 * (exit[1] - entry[1]) * (exit[2] - entry[2])) / 1.0))
```

The interior cells between boundary cells are classified by **winding
number**: if the accumulated winding from boundary crossings is
nonzero, the cell is fully inside (fraction = 1.0). This is the same
winding sweep as approx mode — the only difference is what happens
at boundary cells.

## From dense to sparse

Both fasterize and exactextract produce dense output: a full raster
matrix where every cell has a value (or NA). This is O(area) in memory.

controlledburn produces **sparse output**: only the non-empty cells are
recorded. Interior cells are run-length encoded (`$runs`), boundary
cells have individual records (`$edges`). This is O(perimeter) in
memory.

```{r resolution-scaling, fig.width = 9, fig.height = 3.5, fig.cap = "Resolution scaling: the same polygon at 8×7, 16×14, and 32×28 resolution. Interior cells (blue) grow as area (O(n²)); boundary cells (coral) grow as perimeter (O(n)). Sparse output stores only the blue runs and coral edges — the white cells cost nothing."}
library(geos)
g <- as_geos_geometry(sprintf("POLYGON ((%s))",
  paste(sprintf("%.1f %.1f", poly_x, poly_y), collapse = ", ")))

burns <- lapply(c(8, 16, 32), function(s) {
  r <- burn(g, extent = c(0, nx, 0, ny), dimension = c(as.integer(s), as.integer(s * ny/nx)))
  dx <- nx / s; dy_g <- ny / (s * ny/nx)
  
  cells <- data.frame(
    xmin = numeric(0), xmax = numeric(0),
    ymin = numeric(0), ymax = numeric(0),
    type = character(0)
  )
  
  if (nrow(r$runs) > 0) {
    for (i in seq_len(nrow(r$runs))) {
      row <- r$runs$row[i]
      for (col in r$runs$col_start[i]:r$runs$col_end[i]) {
        cells <- rbind(cells, data.frame(
          xmin = (col - 1) * dx, xmax = col * dx,
          ymin = ny - row * dy_g, ymax = ny - (row - 1) * dy_g,
          type = "interior"
        ))
      }
    }
  }
  
  if (nrow(r$edges) > 0) {
    for (i in seq_len(nrow(r$edges))) {
      row <- r$edges$row[i]; col <- r$edges$col[i]
      cells <- rbind(cells, data.frame(
        xmin = (col - 1) * dx, xmax = col * dx,
        ymin = ny - row * dy_g, ymax = ny - (row - 1) * dy_g,
        type = "boundary"
      ))
    }
  }
  
  cells$panel <- sprintf("%d × %d", s, as.integer(s * ny/nx))
  cells
})

all_cells <- do.call(rbind, burns)
all_cells$panel <- factor(all_cells$panel,
                          levels = c("8 × 7", "16 × 14", "32 × 28"))

ggplot(all_cells) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
                fill = type), colour = "grey80", linewidth = 0.1) +
  geom_polygon(data = data.frame(x = poly_x, y = poly_y),
               aes(x, y), fill = NA, colour = "black", linewidth = 0.6) +
  scale_fill_manual(values = c("interior" = "steelblue", "boundary" = "coral")) +
  facet_wrap(~panel) +
  coord_equal(xlim = c(0, nx), ylim = c(0, ny)) +
  theme_grid() +
  theme(strip.text = element_text(face = "bold"))
```

As resolution increases, the number of interior cells grows as
$$O(n^2)$$ (area) while the number of boundary cells grows as
$$O(n)$$ (perimeter). The sparse representation stores only the
occupied cells — at high resolution, it's overwhelmingly interior
runs, which compress to a handful of `(row, col_start, col_end)`
records per row.

This is why controlledburn can burn 25,954 Antarctic rock outcrop
polygons onto an 8-trillion-cell grid in 8 seconds using 438 MB of
memory. A dense raster would require ~30 TB.

## Summary

|  | fasterize | exactextract | controlledburn |
|--|-----------|-------------|----------------|
| **Algorithm** | Scanline fill, centre rule | Walker + analytical coverage | Both (approx / coverage mode) |
| **Boundary cells** | Binary in/out | Exact fraction | Both options |
| **Output** | Dense raster | Dense raster (or weights) | Sparse tables |
| **Memory** | O(area) | O(area) | O(perimeter) |
| **Primary use** | Raster map production | Zonal statistics | Geometry–grid intersection |
| **Dependencies** | Rcpp, sf, Armadillo | GEOS, Rcpp | cpp11, wk |

The three tools answer the same geometric question. controlledburn
factors the answer into a reusable sparse representation that supports
all three motivations — masking, weighting, and identification — at
scales where dense output is infeasible.

See `vignette("architecture")` for the development history and code
structure.
