---
title: "Rasterizing ice-free Antarctica"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 8)
Sys.setenv("GDAL_DISABLE_READDIR_ON_OPEN" = "EMPTY_DIR")
```

Antarctica's exposed rock — the ice-free ground — has been mapped as
25,954 polygons covering nunataks, coastal outcrops, and mountain
ranges across the continent. The Reference Elevation Model of
Antarctica (REMA) provides a 2m DEM mosaic of the entire ice sheet,
tiled at multiple resolutions.

What does it take to rasterize those polygons onto the REMA grid? At
32m resolution, the grid is 170318 x 182568 pixels. A dense raster at that size would enormous
memory, butcontrolledburn does it in under a second, producing 438 MB of sparse
output, then serves up any individual tile on demand.

## Setup

Three cloud-hosted datasets, read directly via GDAL virtual filesystem
paths. 

```{r data}
# REMA 32m DEM mosaic — we read only the grid spec, no pixel values
dsn <- paste0(
  "/vsicurl/https://raw.githubusercontent.com/mdsumner/rema-ovr/",
  "main/rema-vrt/32m_dem_tiles.vrt")

# Antarctic rock outcrop polygons (v7.3, Burton-Johnson et al.)
shp <- paste0(
  "/vsizip/{/vsicurl/https://github.com/AustralianAntarcticDivision/",
  "rema.proc/raw/refs/heads/master/01_rock_classification/",
  "Medium_resolution_vector_polygons_of_Antarctic_rock_outcrop_-_",
  "VERSION_7.3.zip}/Medium resolution vector polygons of Antarctic ",
  "rock outcrop - VERSION 7.3/",
  "add_rock_outcrop_medium_res_polygon_v7.3.gpkg")

# REMA tile index — for choosing a view window
rema_tiles <- paste0(
  "/vsizip/{/vsicurl/https://data.pgc.umn.edu/elev/dem/setsm/REMA/",
  "indexes/REMA_Mosaic_Index_latest_shp.zip}/REMA_Mosaic_Index_v2_shp")
```

## Read

The raster is never opened — `vapour_raster_info()` reads only the
metadata (extent, dimensions, CRS). The polygons are read as raw WKB
bytes.

```{r read}
library(terra)
tiles <- terra::vect(rema_tiles)

raster_info <- vapour::vapour_raster_info(dsn)
rock_info <- vapour::vapour_layer_info(shp)
rock <- wk::wkb(vapour::vapour_read_geometry(shp),
                 crs = rock_info$projection$Wkt)

cat(sprintf("Grid: %s × %s pixels (%.0f billion cells)\n",
    format(raster_info$dimension[1], big.mark = ","),
    format(raster_info$dimension[2], big.mark = ","),
    prod(as.numeric(raster_info$dimension)) / 1e9))
cat(sprintf("Polygons: %s geometries\n",
    format(length(rock), big.mark = ",")))
```

## Burn the continent

One call rasterizes every polygon onto the full continental grid.
The output is a sparse table of run-length encoded interior cells —
no dense matrix is allocated, no pixel values are touched.

```{r burn}
library(controlledburn)

system.time({
  ba <- burn(rock, extent = raster_info$extent,
             dimension = raster_info$dimension, mode = "approx")
})

ba
cat(sprintf("\nObject size: %.0f MB\n", as.numeric(lobstr::obj_size(ba)) / 1e6))
```

Coverage mode computes exact boundary fractions — slower, but the total
burned area equals the true polygon area to floating-point precision:

```{r burn-coverage}
system.time({
  be <- burn(rock, extent = raster_info$extent,
             dimension = raster_info$dimension, mode = "coverage")
})

be
```

## Look at a tile

The sparse output covers the whole continent but we can extract any
sub-window instantly with `crop_burn()`. Here: a 50 km REMA tile near
Cape Hallett, northern Victoria Land.

```{r tile-locate}
ll <- cbind(lon = 169.6539, lat = -71.50035)
xy <- reproj::reproj_xy(ll, crs(tiles), source = "EPSG:4326")
idx <- which(terra::is.related(tiles, vect(xy), "intersects"))
tile_ex <- as.vector(ext(tiles[idx, ]))
```

```{r tile-approx, fig.cap = "Rock outcrop near Cape Hallett, approx mode (cell-centre rule). White = exposed rock, dark = ice or ocean. Pink outlines are the source polygon boundaries."}
ma <- materialize_chunk(crop_burn(ba, tile_ex))
cat(sprintf("Tile: %d × %d pixels\n", ncol(ma), nrow(ma)))

# Rock polygons that overlap this tile
rock_idx <- unlist(
  geos::geos_strtree_query(
    geos::geos_strtree(geos::as_geos_geometry(rock)),
    tiles[idx, ]))

par(bg = "#1a1a2e", mar = c(3, 3, 2, 1))
plot(tiles[idx, ], col = NA, border = NA,
     main = "Cape Hallett rock outcrop — 32m",
     col.main = "white", col.axis = "grey60")
ximage::ximage(ma, tile_ex, add = TRUE,
               col = c("#1a1a2e", "#e8d5b7"))
plot(rock[rock_idx], add = TRUE, border = "#e05780", lwd = 0.6)
```

```{r tile-coverage, fig.cap = "Same view, coverage mode. Boundary cells carry exact fractional weights — visible as soft edges where the polygon boundary crosses pixel boundaries."}
mc <- materialize_chunk(crop_burn(be, tile_ex))

par(bg = "#1a1a2e", mar = c(3, 3, 2, 1))
plot(tiles[idx, ], col = NA, border = NA,
     main = "Cape Hallett rock outcrop — 32m, exact coverage",
     col.main = "white", col.axis = "grey60")
ximage::ximage(mc, tile_ex, add = TRUE,
               col = hcl.colors(64, "Lajolla"))
plot(rock[rock_idx], add = TRUE, border = "#e05780", lwd = 0.6)
```

At 32m resolution the two modes are nearly identical. The difference
is sub-pixel boundary treatment:

```{r diff}
n_diff <- sum(ma != mc)
cat(sprintf("Cells that differ: %s out of %s (%.3f%%)\n",
    format(n_diff, big.mark = ","),
    format(length(ma), big.mark = ","),
    100 * n_diff / length(ma)))
```

## Summary of the lazy rasterization process

controlledburn burned 25,954 polygons onto a 31-billion-cell grid in
under a second, producing a sparse run-length table small enough to
hold in memory. Any tile can be extracted and materialised on demand.
The raster was never read — only its grid specification. The polygons
were never validated or reprojected — just raw WKB bytes from a
cloud-hosted GeoPackage.

This is rasterization as an O(perimeter) operation: the cost scales
with the complexity of the polygon boundaries, not the number of
pixels in the grid. The sparse output is a reusable intermediate —
downstream processing (zonal statistics, masking, tiling) operates on
the same table without re-burning.

## Creating a raster file

Now we have a very dense version of our rock layer as 
