---
title: "Antarctic rock outcrop at 32m resolution"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
Sys.setenv("GDAL_DISABLE_READDIR_ON_OPEN" = "EMPTY_DIR")
```

controlledburn rasterizes vector geometry into sparse run-length tables
without allocating a dense raster. This example burns 25,954 Antarctic
rock outcrop polygons onto the REMA 32m DEM grid — 851,000 × 913,000
pixels, ~780 billion cells — then extracts a single 50 km tile for
display. The full burn takes under a second.

## Data

Three cloud-hosted sources, read directly via GDAL virtual filesystem
paths:

```{r data}
# REMA 32m DEM mosaic (grid spec only — no pixels read)
dsn <- paste0(
  "/vsicurl/https://raw.githubusercontent.com/mdsumner/rema-ovr/",
  "main/rema-vrt/32m_dem_tiles.vrt")

# Antarctic rock outcrop polygons (v7.3)
shp <- paste0(
  "/vsizip/{/vsicurl/https://github.com/AustralianAntarcticDivision/",
  "rema.proc/raw/refs/heads/master/01_rock_classification/",
  "Medium_resolution_vector_polygons_of_Antarctic_rock_outcrop_-_",
  "VERSION_7.3.zip}/Medium resolution vector polygons of Antarctic ",
  "rock outcrop - VERSION 7.3/",
  "add_rock_outcrop_medium_res_polygon_v7.3.gpkg")

# REMA tile index (for choosing a view window)
rema_tiles <- paste0(
  "/vsizip/{/vsicurl/https://data.pgc.umn.edu/elev/dem/setsm/REMA/",
  "indexes/REMA_Mosaic_Index_latest_shp.zip}/REMA_Mosaic_Index_v2_shp")
```

## Read geometry and grid spec

The raster is never read — only its extent and dimensions. The polygons
are read as WKB.

```{r read}
library(terra)
tiles <- terra::vect(rema_tiles)

raster_info <- vapour::vapour_raster_info(dsn)
rock_info <- vapour::vapour_layer_info(shp)
rock <- wk::wkb(vapour::vapour_read_geometry(shp),
                 crs = rock_info$projection$Wkt)

cat(sprintf("Grid: %d × %d (%.0f billion cells)\n",
    raster_info$dimension[1], raster_info$dimension[2],
    prod(as.numeric(raster_info$dimension)) / 1e9))
cat(sprintf("Polygons: %d geometries\n", length(rock)))
```

## Burn

Both modes burn all 25,954 polygons onto the full continental grid.
Approx mode (cell-centre rule) produces runs only; coverage mode
adds exact boundary fractions.

```{r burn}
library(controlledburn)

system.time({
  ba <- burn(rock, extent = raster_info$extent,
             dimension = raster_info$dimension, mode = "approx")
})

system.time({
  be <- burn(rock, extent = raster_info$extent,
             dimension = raster_info$dimension, mode = "coverage")
})

ba
be
```

## Extract a tile

Pick a location (Cape Hallett area), find the REMA tile that covers it,
crop the sparse burn to that tile, and materialise a dense matrix for
display.

```{r tile, fig.width = 7, fig.height = 7}
ll <- cbind(lon = 169.6539, lat = -71.50035)
xy <- reproj::reproj_xy(ll, crs(tiles), source = "EPSG:4326")
idx <- which(terra::is.related(tiles, vect(xy), "intersects"))

tile_ex <- as.vector(ext(tiles[idx, ]))

# Approx mode tile
ma <- materialize_chunk(crop_burn(ba, tile_ex))

# Find which rock polygons overlap this tile
rock_tile_index <- unlist(
  geos::geos_strtree_query(
    geos::geos_strtree(geos::as_geos_geometry(rock)),
    tiles[idx, ]))

plot(tiles[idx, ])
ximage::ximage(ma, tile_ex, add = TRUE, col = c("black", "white"))
plot(rock[rock_tile_index], add = TRUE, border = "hotpink", lwd = 2)
```

The white cells are rock outcrop at 32m resolution. The pink outlines
are the source polygon boundaries. Each tile materialises to a
`r paste(dim(ma), collapse = " × ")` matrix — small enough to work
with, produced instantly from the sparse burn that covered the entire
continent.

```{r coverage-tile, fig.width = 7, fig.height = 7}
# Coverage mode tile — same view, exact boundary fractions
mc <- materialize_chunk(crop_burn(be, tile_ex))
plot(tiles[idx, ])
ximage::ximage(mc, tile_ex, add = TRUE, col = c("black", "white"))
plot(rock[rock_tile_index], add = TRUE, border = "hotpink", lwd = 2)
```

At 32m resolution the two modes are nearly indistinguishable. The
difference is sub-pixel boundary fractions:

```{r diff}
cat(sprintf("Cells that differ: %d out of %d (%.4f%%)\n",
    sum(ma != mc), length(ma),
    100 * sum(ma != mc) / length(ma)))
```
